Efficient pedigree recording for fast population genetics simulation

### 2018-06-06 review-responses.tex

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We'd like to thank the reviewers for a careful reading of the paper;
as detailed below, their helpful comments have resulted in a substantial amount of clarification
about the overall structure and purpose of the paper.
We hope you agree that our changes have addressed their comments and concerns.

There is one change of note that we should explain:
since initial submission, we have decided to separate out the tools for dealing with tree sequences
into a standalone package, called \tskit,
from the coalescent simulation code (called \msprime).
This should help avoid confusion.
However, although we're a good ways towards this goal,
we haven't yet completed the clean separation,
so that currently the \tskit{} page instructs users to install \msprime{}.
We think this slightly confusing situation is preferable to the alternative,
which would have the terms in our paper outdated in a few months
(when we move the tools discussed here
from \msprime{} to \tskit{}).

@@

\reply{
    As the reviewer suggests, our main goal here is to provide easy access
    to the \emph{algorithm}, not to a forward simulation model --
    we aren't describing a forward simulation model, 
    but rather a way to speed up and get more information out of an existing one.
    But, is the algorithm accessible?
    An interface to our algorithms is available at the python level or at the C level.
    The python interface is documented through the \msprime{} documentation
    (but you need to select ``latest'': \url{https://msprime.readthedocs.io/en/latest/});
    and closely mirrors the underlying C library.
    We have also provided a link to the  developing set of tutorials 
    at \url{https://github.com/tskit-dev/tutorials} \llname{ll:tutorials}.
}

\begin{point}{}
Figure 1 shows that pedigree recording seems to scale poorly in N. Going from
N=1e3 to N=1e4 results in what seems like a $\sim 5 \times$ increase in runtime, but going
up to N=5e4 results in a $> 10 \times$ increase in runtime. Given your attention to
scaling, is this expected?
\end{point}

\reply{
    Good spot: this is actually the fault of a quadratic algorithm for counting mutations
    under the hood in \fwdpp{}.
    In fact, some of us (KRT) have used ideas from tree sequences to fix this bottleneck,
    and the fix will be part of the next \fwdpp{}.
}

### 2018-06-06 forwards_paper.tex
msprime -> tskit

\begin{abstract}
    In this paper we describe how to efficiently record the entire genetic history of a population
    in forwards-time, individual-based population genetics simulations with arbitrary breeding models,
    population structure and demography.
    This approach dramatically reduces the computational burden of tracking individual genomes
    by allowing us to simulate only those loci that may affect reproduction
    (those having non-neutral variants).
    The genetic history of the population is recorded as a succinct tree sequence
    as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    Recording the results of each breeding event
    results in a linear growth in the amount of information that must be stored.
    However, there is a great deal of redundancy in this information,
    since the majority of individuals simulated over the course of the simulation
    will leave no descendents in the eventual sample.
    We solve this storage problem by providing an algorithm to quickly `simplify'
    a tree sequence by removing this irrelevant history for a given set of genomes.
    By periodically simplifying the history with respect to the extant population,
    we show that the total storage space required is modest
    and overall large efficiency gains can be made over classical forward-time simulations.
    We implement a general-purpose framework for recording and simplifying genealogical data,
    which can be used to make simulations of any population model more efficient.
    We modify two popular forwards-time simulation frameworks to use this new approach
    and observe efficiency gains in large, whole-genome simulations of one to two orders of magnitude.
    In addition to speed, our method for recording pedigrees has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
        can be stored in $O(N \log N + M)$ space,
        making it feasible to store a simulation's entire final generation
        (and its history) rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation 
        of deep history.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees}
and thus bringing within-population trees into better focus;
second,
producing analytical methods to estimate parameters of interest from genetic data; \revpoint{1}{5}
and finally, providing a computationally feasible method
to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
under the following important assumptions: \revpoint{2}{3}
(a) random mating,
(b) neutrality,
(c) large population size,
and (d) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent \citep{hudson1990gene, Neuhauser1997-nn},
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical
space~\citep{barton2010new,kelleher2014coalescent}
and/or have large numbers of loci under various sorts of selection.
Usually, the relationship between the life history of a species --
fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
More mechanistic models are possible using ``forwards--backwards'' simulations,
that first simulate population size changes forwards in time
and then thread a coalescent backwards \citep{ray2010splatche2}, \revpoint{1}{3}
but these still require the assumptions above to be met for each subpopulation.
The last assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples approaching the size of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
\fwdpp{} \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\sim 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

@@

In this paper, we describe a storage method for \emph{succinct tree sequences}
(and hence, genome sequence) as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its space usage is close to optimal,
while still allowing efficient retrieval of information (see, e.g., \citet{gog2014theory}).
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide 
in principle much more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
These marginal genealogies enable fast data storage and processing,
but also provide additional information that can be used to better understand
the notoriously complex dynamics of population genetics. \revpoint{1}{4}
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.
This work originated as improvements to the algorithmic tools and data structures 
in the coalescent simulator \msprime{}; \linelabel{msprime_rel}
the tools are currently available bundled with the Python package \msprime{},
but will soon be separately available as a package named \tskit{}, 
which will include a Python API and an embeddable C library.

@@

The strategy described above is only of interest if it is computationally feasible.
Therefore,
we begin by benchmarking the performance improvement achieved by this method,
implemented using the forwards-time simulation library \fwdpp{} \citep{fwdpp}
and tree sequence tools implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences efficiently in several \emph{tables};
(b) an algorithm to record these tables during a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
Finally, we analyze the run time and space complexity of our general-purpose method.

@@

% When tracking neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for these simulations).
Pedigree tracking dramatically reduced runtimes, as shown in Figure~\ref{fig:runtimes_selection},
producing a relative speedup of up to $\approx 50$ fold relative to standard simulations that track neutral mutations
(Figure~\ref{fig:relative_speedup_selection}).
Pedigree tracking results in greater relative speedups for larger $N$
and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).
Importantly, runtimes are approximately linear in region size $\rho$ when pedigree tracking
(partially obscured by the log scale of the horizontal axis in Figure~\ref{fig:runtimes_selection}).
%% don't know what an 'edge' is yet
% The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
% each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative behavior,
and a numerically larger speedup by using pedigree tracking (see Appendix~\ref{ss:timing_nosel}).

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The leftmost panels show the tree sequence pictorially in two different ways:
        (top) a sequence of tree topologies; the first tree extends from genomic position 0 to 5,
        and the second from 5 to 10; and
        (bottom) the edges that define these topologies,
        displayed over their corresponding genomic segment
        (for instance, the edge from node 2 to node 4 is present only on the interval from 0 to 5).
        The remaining panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations)
        defined by \msprime.
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

This encoding of a sequence of trees and accompanying mutational information is
very concise. To illustrate this, we used \msprime{} to simulate 500,000 samples of a
$200$ megabase chromosome with human-like parameters: $N_e=10^4$ and per-base mutation and
recombination rates of $10^{-8}$ per generation. This resulted
in about 1 million distinct marginal trees and $1.1$ million infinite-sites
mutations. The HDF5 file encoding the node, edge, site and mutation tables (as
described above) for this simulation consumed 157MiB of storage space. Using
the \msprime\ Python API, the time required to load this file into memory was
around 1.5 seconds, and the time required to iterate over all 1 million trees
was 2.7 seconds. In contrast, recording the topological information in Newick
format would require around 20 TiB and storing the genotype information
in VCF would require about 1 TiB (giving a compression factor of 144,000 in
this instance).
Working with either the Newick or VCF encoding
of this dataset would likely require several
days of CPU time just to read the information into memory.

\paragraph{Validity of a set of tables}
Given a set of node and edge tables as described above,
there are only two requirements that ensure the tables
describe a valid tree sequence.
% (There are essentially no such requirements on the site and mutation tables.)
These are:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
A pair of node and edge tables that satisfy these two requirements
is guaranteed to uniquely describe at each point on the genome
a collection of directed, acyclic graphs -- in other words, a forest of trees.
For some applications it is necessary to check that at every point
there is only a \emph{single} tree.
Checking this is more difficult, but is implemented in \msprime{}'s API.
For efficiency, \msprime{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tables emitted by a forwards-time simulation.
\msprime{}'s API includes tools to rectify this by first sorting % (using \texttt{sort\_tables})
and then using the \texttt{simplify} algorithm described below, which works on sorted tables
and is guaranteed to produce a valid, \msprime{}-ready tree sequence.

%%%%%%
\subsection*{The \msprime\ Tables API}

The facilities for working with succinct tree sequences are implemented as part
of the \msprime\ Python API, which provides a powerful platform for processing
tree topology and mutation data. The new portions of \msprime{} that we discuss
here are dedicated to tree sequence input and output using simple tables of
data, as described above, so we refer to this as the ``Tables API''.

@@

The \msprime{} Python Tables API provides a general purpose toolkit for importing
and processing succinct tree sequences, and a collection of tutorials are being developed
at \url{https://github.com/tskit-dev/tutorials}. \linelabel{ll:tutorials}
Interoperation with Python simulators is then straightforward.  
The implementation we benchmark here uses \texttt{pybind11} 
(\url{https://github.com/pybind/pybind11/}) to interface with the \fwdpp{} \cpp{} API \citep{fwdpp}. 
No modifications were
required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,
and perform simple processing of the recombination breakpoints from each mating
event to generate node and edge data. This information is then periodically copied
to the \msprime\ Tables API, where it is sorted and simplified.

\paragraph{Flexibility.}
To demonstrate the flexibility provided by the Tables API and provide an
implementation that decouples forward simulation internals from transfer of data
to \msprime, we also implemented a version of the simulations described in
``Simulation benchmarks'' separately in Python, described in Appendix~\ref{ss:simupop}.
In this proof-of-concept implementation, the simulation engine (we use \simupop{}, \citet{simupop})
invokes callbacks at critical points of the simulation, and we infer nodes and edges
from the information that is provided. Rows are appended to the tables
one-by-one, and the tables are periodically sorted and simplified to control
memory usage.
Benchmarking results from this implementation are shown (alongside results from \fwdpp{})
for simulations without selection in Appendix~\ref{ss:timing_nosel}:
a relatively modest speedup of around $5 \times$ is achieved, likely due to increased overhead.

@@

\paragraph{Essential steps in tree sequence recording}
Since a tree sequence can record the history of genetic inheritance in any situation
(requiring only unambiguous inheritance and no time travel),
any individual-based population genetics simulator can maintain a tree sequence
with only a little bookkeeping. \revpoint{1}{1}
We have furthermore provided several tools to minimize this bookkeeping,
only requiring a little one-way output at the birth of each new individual.
Concretely, to record a tree sequence, including mutations,
a simulator must record for each new genome
(so, twice for each new diploid individual): \revpoint{1}{6}
\begin{itemize}
    \item the birth time of the genome in the Node Table,
    \item the segments the genome inherits from its parental genomes in the Edge Table,
    \item the locations of any new mutations in the Site Table,
    \item and the derived state of these new mutations in the Mutation Table
        (as well as the identiy of this genome the mutations appeared in).
\end{itemize}
Each of these can be simply appended to the ends of the respective tables.
Besides this, simplification should be run every once in a while (e.g., every 100 generations).
Before simplification, time in the Node Table must be translated to ``time ago'' if it is not already.
(This was avoided in Algorithm~\algref{W} since there $t$ denoted
``time until the end of the simulation''.)
There are also several ``cleaning'' steps, for which we provide functions in the Tables API:
\emph{sorting} according to several criteria for algorithmic efficiency;
and removing any duplicate sites from the Site Table.
After simplification, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \msprime{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.

@@

The tree sequences produced by default by this method
are very compact, storing genotype \emph{and} genealogical information
in a small fraction of the space taken by a compressed VCF file.
The format also allows highly efficient processing for downstream analysis.
Efficient processing is possible because many statistics of interest for population genetics
are naturally expressed in terms of tree topologies,
and so can be quickly computed from the trees underlying the tree sequence format.
For example, pairwise nucleotide diversity $\pi$, is the average density of
differences between sequences in the sample.
To compute this directly from sequence data at $m$ sites in $n$ samples
requires computing allele frequencies, taking $O(nm)$ operations.
By using the locations of the mutations on the marginal trees,
and the fact that these are correlated,
sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
can do this in roughly $O(n + m + t \log n)$ operations, where
$t$ is the number of distinct trees.
The \msprime{} API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, which took about 0.7 seconds when
applied to an example simulation of 100 megabases of human-like
sequence for 200,000 samples (about 500K sites). The corresponding
numeric genotype matrix required about 95GiB of RAM, and
calculating $\pi$ took about 66 seconds with NumPy.

@@

In this article, we applied our methods for storing trees to the problem of pedigree recording in a forward-time
simulation.  However, the method applies to any simulation scheme generating nodes and edges.  For example, one could
use the methods described here to generate succinct tree sequences under coalescent processes not currently implemented
in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc}, using the structured coalescent to
model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn},
or the coalescent within a known pedigree.  For such models, one
could in principle generate tables of nodes and edges to be simplified in \msprime{}.  The resulting succinct tree
sequence object would be in the same format as those generated by \msprime{}'s built-in simulate function, and therefore
compatible with existing methods for downstream analyses.

@@

% JK: This needs more context. Probably not worth putting in here.
% \paragraph{Next steps.}
% While our methods result in significant performance
% gains over current simulation methods, they are far from optimal. Because
% forwards time simulation is highly CPU intensive, most simulation engines
% are written in \justc{} or \cpp{} out of necessity. Therefore, the simplest and most
% efficient approach would be to call the \msprime{} \justc{} library directly as part of
% the core simulation logic, in the manner outlined in Algorithm \algref{W}. If designed
% in this way, the simulation engine would not need to keep copies of the node and edge information.
% To facilitate this usage, we plan to release a standalone, embeddable \justc{} library encompassing the core
% tree sequence functionality, which should make recording and
% outputting tree sequences from any \justc{}/\cpp{} code straightforward.

@@

We implemented simulations and the connection to \msprime{} in \cpp{}, using \fwdpp{} library functions and
interface code using a continuum-sites model for both mutation
and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
on \fwdpp{} (version 0.5.7).  The majority of results are presented based on a single-threaded implementation.  However,
we also implemented a parallelized version using Python's \texttt{queue.Queue} to run the simplification step in a separate Python
thread. Our implementation allows a maximum of four simplification intervals to be in the queue at once. This
parallelized version also performed fitness calculation in parallel using two threads of execution in C++.

@@

% \paragraph{Requirements.}
% The simulations as implemented depend on the class \texttt{RecombCollector} provided by
% the Python package \ftprime{} 0.0.6rc0,
% available at \url{https://www.github.com/ashander/ftprime}.
% Its requirements are encoded in an \texttt{environment.yml} file, compatible
% with conda, but \msprime{} must be updated to the version listed below via pip
% from GitHub \jda{TODO add command?}.
%
% In brief the requirements are:
%
% \begin{enumerate}
%     \item \msprime{} 0.5.0
%     \item \simupop{} 1.1.8.3
%     \item python 3.5
% \end{enumerate}

>>>>

\begin{abstract}
    In this paper we describe how to efficiently record the entire genetic history of a population
    in forwards-time, individual-based population genetics simulations with arbitrary breeding models,
    population structure and demography.
    This approach dramatically reduces the computational burden of tracking individual genomes
    by allowing us to simulate only those loci that may affect reproduction
    (those having non-neutral variants).
    The genetic history of the population is recorded as a succinct tree sequence
    as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    Recording the results of each breeding event
    results in a linear growth in the amount of information that must be stored.
    However, there is a great deal of redundancy in this information,
    since the majority of individuals simulated over the course of the simulation
    will leave no descendents in the eventual sample.
    We solve this storage problem by providing an algorithm to quickly `simplify'
    a tree sequence by removing this irrelevant history for a given set of genomes.
    By periodically simplifying the history with respect to the extant population,
    we show that the total storage space required is modest
    and overall large efficiency gains can be made over classical forward-time simulations.
    We implement a general-purpose framework for recording and simplifying genealogical data,
    which can be used to make simulations of any population model more efficient.
    We modify two popular forwards-time simulation frameworks to use this new approach
    and observe efficiency gains in large, whole-genome simulations of one to two orders of magnitude.
    In addition to speed, our method for recording pedigrees has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
        can be stored in $O(N \log N + M)$ space,
        making it feasible to store a simulation's entire final generation
        (and its history) rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation 
        of deep history.
    The software for recording and processing tree sequences is named \tskit.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees}
and thus bringing within-population trees into better focus;
second,
producing analytical methods to estimate parameters of interest from genetic data; \revpoint{1}{5}
and finally, providing a computationally feasible method
to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
under the following important assumptions: \revpoint{2}{3}
(a) random mating,
(b) neutrality,
(c) large population size,
and (d) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent \citep{hudson1990gene, Neuhauser1997-nn},
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical
space~\citep{barton2010new,kelleher2014coalescent}
and/or have large numbers of loci under various sorts of selection.
Usually, the relationship between the life history of a species --
fecundity and mortality schedules, density-dependent effects on fitness, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
More mechanistic models are possible using ``forwards--backwards'' simulations,
that first simulate population size changes forwards in time
and then thread a coalescent backwards \citep{ray2010splatche2}, \revpoint{1}{3}
but these still require the assumptions above to be met for each subpopulation.
The last assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples approaching the size of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
\fwdpp{} \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to around $5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

@@

In this paper, we describe a storage method for \emph{succinct tree sequences}
(and hence, genome sequence) as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its space usage is close to optimal,
while still allowing efficient retrieval of information (see, e.g., \citet{gog2014theory}).
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide 
in principle much more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
These marginal genealogies enable fast data storage and processing,
but also provide additional information that can be used to better understand
the notoriously complex dynamics of population genetics. \revpoint{1}{4}
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.
This work originated as improvements to the algorithmic tools and data structures 
in the coalescent simulator \msprime{}; \linelabel{msprime_rel}
the software tools described here for working with tree sequences are referred to as \tskit{};
they are currently bundled with the Python package \msprime{},
but will soon be separately available as a Python API and an embeddable C library.

@@

The strategy described above is only of interest if it is computationally feasible.
Therefore,
we begin by benchmarking the performance improvement achieved by this method,
implemented using the forwards-time simulation library \fwdpp{} \citep{fwdpp}
and tree sequence tools implemented in \tskit{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \tskit{} Python API,
for recording tree sequences efficiently in several \emph{tables};
(b) an algorithm to record these tables during a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
Finally, we analyze the run time and space complexity of our general-purpose method.

@@

% When tracking neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for these simulations).
Pedigree tracking dramatically reduced runtimes, as shown in Figure~\ref{fig:runtimes_selection},
producing a relative speedup of up to around 50 fold relative to standard simulations that track neutral mutations
(Figure~\ref{fig:relative_speedup_selection}).
Pedigree tracking results in greater relative speedups for larger $N$
and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).
Importantly, runtimes are approximately linear in region size $\rho$ when pedigree tracking
(partially obscured by the log scale of the horizontal axis in Figure~\ref{fig:runtimes_selection}).
%% don't know what an 'edge' is yet
% The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
% each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative behavior,
and a numerically larger speedup by using pedigree tracking (see Appendix~\ref{ss:timing_nosel}).

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The leftmost panels show the tree sequence pictorially in two different ways:
        (top) a sequence of tree topologies; the first tree extends from genomic position 0 to 5,
        and the second from 5 to 10; and
        (bottom) the edges that define these topologies,
        displayed over their corresponding genomic segment
        (for instance, the edge from node 2 to node 4 is present only on the interval from 0 to 5).
        The remaining panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations).
        \label{fig:example_tree_sequence}
    }
\end{figure}

@@

This encoding of a sequence of trees and accompanying mutational information is
very concise. To illustrate this, we used \msprime{} to simulate 500,000 samples of a
$200$ megabase chromosome with human-like parameters: $N_e=10^4$ and per-base mutation and
recombination rates of $10^{-8}$ per generation. This resulted
in about 1 million distinct marginal trees and $1.1$ million infinite-sites
mutations. The HDF5 file encoding the node, edge, site and mutation tables (as
described above) for this simulation consumed 157MiB of storage space. Using
the \tskit{} Python API, the time required to load this file into memory was
around 1.5 seconds, and the time required to iterate over all 1 million trees
was 2.7 seconds. In contrast, recording the topological information in Newick
format would require around 20 TiB and storing the genotype information
in VCF would require about 1 TiB (giving a compression factor of 144,000 in
this instance).
Working with either the Newick or VCF encoding
of this dataset would likely require several
days of CPU time just to read the information into memory.

\paragraph{Validity of a set of tables}
Given a set of node and edge tables as described above,
there are only two requirements that ensure the tables
describe a valid tree sequence.
% (There are essentially no such requirements on the site and mutation tables.)
These are:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
A pair of node and edge tables that satisfy these two requirements
is guaranteed to uniquely describe at each point on the genome
a collection of directed, acyclic graphs -- in other words, a forest of trees.
For some applications it is necessary to check that at every point
there is only a \emph{single} tree.
Checking this is more difficult, but is implemented in \tskit{}'s API.
For efficiency, \tskit{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tables emitted by a forwards-time simulation.
\tskit{}'s API includes tools to rectify this by first sorting % (using \texttt{sort\_tables})
and then using the \texttt{simplify} algorithm described below, which works on sorted tables
and is guaranteed to produce a valid, \tskit{}-ready tree sequence.

%%%%%%
\subsection*{The \tskit{} Tables API}

The facilities for working with succinct tree sequences are implemented as part
of the \tskit{} Python API, which provides a powerful platform for processing
tree topology and mutation data. The portions of \tskit{} that we discuss
here are dedicated to tree sequence input and output using simple tables of
data, as described above, so we refer to this as the ``Tables API''.

@@

The \tskit{} Python Tables API provides a general purpose toolkit for importing
and processing succinct tree sequences, and a collection of tutorials are being developed
at \url{https://github.com/tskit-dev/tutorials}. \linelabel{ll:tutorials}
Interoperation with Python simulators is then straightforward.  
The implementation we benchmark here uses \texttt{pybind11} 
(\url{https://github.com/pybind/pybind11/}) to interface with the \fwdpp{} \cpp{} API \citep{fwdpp}. 
No modifications were
required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,
and perform simple processing of the recombination breakpoints from each mating
event to generate node and edge data. This information is then periodically copied
to the \tskit{} Tables API, where it is sorted and simplified.

\paragraph{Flexibility.}
To demonstrate the flexibility provided by the Tables API and provide an
implementation that decouples forward simulation internals from transfer of data
to \tskit, we also implemented a version of the simulations described in
``Simulation benchmarks'' separately in Python, described in Appendix~\ref{ss:simupop}.
In this proof-of-concept implementation, the simulation engine (we use \simupop{}, \citet{simupop})
invokes callbacks at critical points of the simulation, and we infer nodes and edges
from the information that is provided. Rows are appended to the tables
one-by-one, and the tables are periodically sorted and simplified to control
memory usage.
Benchmarking results from this implementation are shown (alongside results from \fwdpp{})
for simulations without selection in Appendix~\ref{ss:timing_nosel}:
a relatively modest speedup of around $5 \times$ is achieved, likely due to increased overhead.

@@

\paragraph{Essential steps in tree sequence recording}
Since a tree sequence can record the history of genetic inheritance in any situation
(requiring only unambiguous inheritance and no time travel),
any individual-based population genetics simulator can maintain a tree sequence
with only a little bookkeeping. \revpoint{1}{1}
We have furthermore provided several tools to minimize this bookkeeping,
only requiring a little one-way output at the birth of each new individual.
Concretely, to record a tree sequence, including mutations,
a simulator must record for each new genome
(so, twice for each new diploid individual): \revpoint{1}{6}
\begin{itemize}
    \item the birth time of the genome in the Node Table,
    \item the segments the genome inherits from its parental genomes in the Edge Table,
    \item the locations of any new mutations in the Site Table,
    \item and the derived state of these new mutations in the Mutation Table
        (as well as the identiy of this genome the mutations appeared in).
\end{itemize}
Each of these can be simply appended to the ends of the respective tables.
Besides this, simplification should be run every once in a while (e.g., every 100 generations).
Before simplification, time in the Node Table must be translated to ``time ago'' if it is not already.
(This was avoided in Algorithm~\algref{W} since there $t$ denoted
``time until the end of the simulation''.)
There are also several ``cleaning'' steps, for which we provide functions in the Tables API:
\emph{sorting} according to several criteria for algorithmic efficiency;
and removing any duplicate sites from the Site Table.
After simplification, since $\tsimplify{P', \Nt, \Et}$ results in the $N$ individuals in $P'$
being relabeled in \tskit{}'s tables as $0, 1, \ldots, N-1$,
there must be bookkeeping that keeps in sync the individual IDs as recorded by the simulator
with the node IDs recorded in the tables.

@@

The tree sequences produced by default by this method
are very compact, storing genotype \emph{and} genealogical information
in a small fraction of the space taken by a compressed VCF file.
The format also allows highly efficient processing for downstream analysis.
Efficient processing is possible because many statistics of interest for population genetics
are naturally expressed in terms of tree topologies,
and so can be quickly computed from the trees underlying the tree sequence format.
For example, pairwise nucleotide diversity $\pi$, is the average density of
differences between sequences in the sample.
To compute this directly from sequence data at $m$ sites in $n$ samples
requires computing allele frequencies, taking $O(nm)$ operations.
By using the locations of the mutations on the marginal trees,
and the fact that these are correlated,
sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
can do this in roughly $O(n + m + t \log n)$ operations, where
$t$ is the number of distinct trees.
The \tskit{} API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, which took about 0.7 seconds when
applied to an example simulation of 100 megabases of human-like
sequence for 200,000 samples (about 500K sites). The corresponding
numeric genotype matrix required about 95GiB of RAM, and
calculating $\pi$ took about 66 seconds with NumPy.

@@

In this article, we applied our methods for storing trees to the problem of pedigree recording in a forward-time
simulation.  However, the method applies to any simulation scheme generating nodes and edges.  For example, one could
use the methods described here to generate succinct tree sequences under coalescent processes not currently implemented
in \msprime{}, such as the coalescent with gene conversion \citep{Wiuf2000-rc}, using the structured coalescent to
model various forms of natural selection \citep{Kaplan1988-in,Kaplan1989-rt,Braverman1995-gn},
or the coalescent within a known pedigree.  For such models, one
could in principle generate tables of nodes and edges to be simplified in \tskit{}.  The resulting succinct tree
sequence object would be in the same format as those generated by \tskit{}'s built-in simulate function, and therefore
compatible with existing methods for downstream analyses.

@@

% JK: This needs more context. Probably not worth putting in here.
% \paragraph{Next steps.}
% While our methods result in significant performance
% gains over current simulation methods, they are far from optimal. Because
% forwards time simulation is highly CPU intensive, most simulation engines
% are written in \justc{} or \cpp{} out of necessity. Therefore, the simplest and most
% efficient approach would be to call the \tskit{} \justc{} library directly as part of
% the core simulation logic, in the manner outlined in Algorithm \algref{W}. If designed
% in this way, the simulation engine would not need to keep copies of the node and edge information.
% To facilitate this usage, we plan to release a standalone, embeddable \justc{} library encompassing the core
% tree sequence functionality, which should make recording and
% outputting tree sequences from any \justc{}/\cpp{} code straightforward.

@@

We implemented simulations and the connection to \tskit{} in \cpp{}, using \fwdpp{} library functions and
interface code using a continuum-sites model for both mutation
and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
on \fwdpp{} (version 0.5.7).  The majority of results are presented based on a single-threaded implementation.  However,
we also implemented a parallelized version using Python's \texttt{queue.Queue} to run the simplification step in a separate Python
thread. Our implementation allows a maximum of four simplification intervals to be in the queue at once. This
parallelized version also performed fitness calculation in parallel using two threads of execution in C++.

@@

% \paragraph{Requirements.}
% The simulations as implemented depend on the class \texttt{RecombCollector} provided by
% the Python package \ftprime{} 0.0.6rc0,
% available at \url{https://www.github.com/ashander/ftprime}.
% Its requirements are encoded in an \texttt{environment.yml} file, compatible
% with conda, but \tskit{} must be updated to the version listed below via pip
% from GitHub \jda{TODO add command?}.
%
% In brief the requirements are:
%
% \begin{enumerate}
%     \item \msprime{} 0.5.0
%     \item \simupop{} 1.1.8.3
%     \item python 3.5
% \end{enumerate}
